A Parallel Algorithm for Solving Global Optimization Problems and Its Application for Tuning Hyperparameters of AI Methods

### 2025-04-11 2025/RuSCDays/RSD-2025.tex
pred-final version

%(рус.) Данная статья продолжает развитие подхода к параллельному решению задач липшицевой глобальной оптимизации, предложенного в работах [наши работы]. Новым элементом является сочетание способов распараллеливания и предложенной в [Усова] схемы работы с частично вычислимыми функциями. Основная часть публикации имеет следующую структуру. В Section 2 приведена математическая постановка решаемой задачи. Section 3 содержит описание параллельного алгоритма глобального поиска. В Section 4 описан способ решения задач со смешанными (дискретными и непрерывными) параметрами, а также обсуждается способ работы с функциями, определенными не во всей области поиска. Section 5 содержит результаты вычислительных экспериментов, демонстрирующих эффективность предлагаемого подхода при решении задачи оптимизации гиперпараметров методов прогнозирования значений временного ряда.
This paper continues to develop the approach to solving Lipschitz global optimization problems in parallel as proposed in [REFS ON OUR WORKS]. The new element is the combination of parallelization methods and the scheme for handling partially computable functions proposed in \cite{Usova2024}. The main part of the publication has the following structure.  Section \ref{sec:PS} presents the mathematical formulation of the problem to be solved. Section \ref{sec:PA} contains a description of the parallel global search algorithm. Section \ref{sec:MP} describes the method for solving problems with mixed (discrete and continuous) parameters, and also discusses the method for handling functions that are not defined in the entire search domain. Section \ref{sec:RCE} gives the results of computational experiments demonstrating the effectiveness of the proposed approach in solving the problem of optimizing the hyperparameters of time series forecasting methods.

@@

%(рус.) 6. Вычислить точку проведения нового испытаний x^{k+p+1} ∈ (x_(t-1),x_t ) в соответствии с формулами
\item Calculate the point of the new trial $x^{k+p+1} \in (x_{t-1}, x_t)$ according to the formulas
\begin{equation}\label{new_point_1} 
x^{k+p+1} = \frac{x_t+x_{t-1}}{2}, \text{ if } v(x_{t-1}) \neq v(x_t),
\end{equation}
\begin{equation}\label{new_point_2} 
x^{k+p+1} = \frac{x_t+x_{t-1}}{2}-\textit{sign}{(z_t-z_{t-1})}\frac{|z_t-z_{t-1}|^N}{2r\mu^N}, \text{ if } v(x_{t-1})=v(x_t).
\end{equation}
%(рус.) Сразу после вычисления точки очередного испытания процесс-мастер добавляет ее в множество S и пересылает процессу-рабочему, который начинает проведение испытания в ней.
Immediately after calculating the point of the next trial, the master process adds it to the set $S$ and forwards it to the worker process, which starts a trial there.
\end{enumerate}

@@

%(рус.) Также требуется дополнить правило выбора точки нового испытания для обработки ситуаций, когда обе граничные точки имеют нулевой индекс:
It is also necessary to supplement the rule for selecting a new trial point to handle situations where both boundary points have a zero index:
\begin{equation}\label{new_point_1_2} 
x^{k+p+1} = \frac{x_t+x_{t-1}}{2}, \text{ if } v(x_{t-1}) \neq v(x_t) \text{ or } v(x_{t-1})=v(x_t)=0,
\end{equation}
\begin{equation}\label{new_point_2_2} 
x^{k+p+1} = \frac{x_t+x_{t-1}}{2}-\textit{sign}{(z_t-z_{t-1})}\frac{|z_t-z_{t-1}|^N}{2r\mu^N}, \text{ if } v(x_{t-1})=v(x_t) = 1.
\end{equation}

@@

%(рус.) Рассмотренные в статье методы реализованы в iOpt parallel solver, разрабатываемом в ННГУ им. Н.И. Лобачевского и предназначенном для решения многомерных задач глобальной оптимизации. 
The methods considered in this paper are implemented in the iOpt parallel solver \cite{iOptURL} +REF!!!!!, which is being developed at Lobachevsky State University of Nizhni Novgorod  for solving multidimensional global optimization problems.

%(рус.) В качестве модельной задачи нами была использована задача настройки гиперпараметров методов предсказания значений временного ряда (Time Series Forecasting). В качестве исходных данных был использован датасет montly beer prodaction [5.1], содержащий 476 valid data points. 
We used the task of hyperparameter tuning for Time Series Forecasting methods as a model problem. The monthly beer production dataset \cite{MonthlyBeerDataset,MonthlyBeerArticle} containing 476 valid data points was used as input data.

@@

%(рус.) Построенный с использованием FEDOT pipeline включает в себя комбинацию из двух преобразований: lagged и cgru. Lagged-преобразование выполняется путем взятия значения переменной в предыдущий момент времени и включения его в модель в качестве признака в текущий момент времени. Для этого данные временного ряда сдвигаются на определенное количество шагов, которые называются лагом или временной задержкой. В качестве используемой нейронной сети будет выступать рекуррентная сеть (GRU), использующая свертку (CGRU).
The pipeline constructed using FEDOT involves a combination of two transformations: lagged and cgru, Fig.~\ref{fig2}. Lagged transformation is performed by taking the value of a variable at a previous point in time and incorporating it into the model as a feature at the current point in time. For this purpose, the time series data is shifted by a certain number of steps, which are called  a lag  or time delay. A recurrent network (GRU) using convolution (CGRU) will serve as the neural network.

@@

%(рус.) У cgru варьировались 4 дискретных параметра: cnn1_kernel_size  {4, 5}, cnn2_kernel_size  {5, 6}, cnn1_output_size  {32, 64}, cnn2_output_size  {32, 64}; и 2 непрерывных параметра: hidden_size  [20, 200], learning_rate  [0.0005, 0.005]. У остальных параметров были зафиксированы default values: batch_size = 64, num_epochs = 50, optimizer = 'adamw', loss = ‘mse’.
For cgru, 4 discrete and 2 continuous parameters were varied:
\begin{itemize}
\item cnn1\_kernel\_size $\in \{4, 5\}$,
\item cnn2\_kernel\_size $\in \{5, 6\}$, 
\item cnn1\_output\_size $\in \{32, 64\}$, 
\item cnn2\_output\_size $\in \{32, 64\}$;
\item hidden\_size $\in [20, 200]$, 
\item learning\_rate $\in [0.0005, 0.005]$.
\end{itemize}

@@

%(рус.) Computations were conducted on the supercomputer “Lobachevsky” of the Nizhny Novgorod State University (operating system CentOS 7, control system SLURM, each supercomputer node had two processors Intel Xeon Silver 4310T CPU@2.30GHz (20 cores), 64Gb RAM). were used. Computations were performed using Python 3.9, FEDOT 0.7.5 [5.2, 5.2.1], iOpt  [20] and Optuna [5.3, 5.3.1].
Calculations were performed on the ``Lobachevsky'' supercomputer of the Nizhni Novgorod State University (CentOS 7 operating system, SLURM control system, each supercomputer node had two Intel Xeon Silver 4310T processors CPU@2.30GHz (20 cores), 64Gb RAM). Data analyses were carried out using Python 3.9, FEDOT 0.7.5 \cite{FEDOT}, iOpt \cite{iOptURL} and Optuna \cite{OptunaURL}.

@@

\begin{table}[]
\centering
\caption{Results of solving the problem using iOpt and Optuna.}\label{tab1}
\begin{tabular}{|l|c|c|c|c|}
\hline
\multirow{2}{*}{P} & \multicolumn{2}{c|}{iOpt} & \multicolumn{2}{c|}{Optuna} \\
\cline{2-5} 
 & Solving time & Final metric & Solving time & Final metric \\
\hline
1 & 6574,28 & 13,56 & 9554,08 & 13,43 \\
2 & 3579,27 & 13,97 & 5127,50 & 13,69 \\
4 & 1998,70 & 13,62 & 2598,68 & 13,26 \\
8 & 1331,79 & 13,20 & 1275,45 & 13,26 \\
16 & 846,69 & 13,31 & 1148,52 & 13,39 \\
20 & 699,94 & 13,98 & 1578,67 & 13,32 \\
\hline
\end{tabular}
\end{table}

%(рус.) Ключевой характеристикой параллельных алгоритмов является ускорение. Мы будем использовать традиционное понятие ускорения как отношение времени последовательного запуска ко времени решения задачи в параллельном режиме. На Fig.3 приведены графики ускорения, полученного при параллельном запуске iOpt и OPTUNA. Наглядно видно, что iOpt на большом числе процессов показывает лучшее ускорение по сравнению с OPTUNA.
The key characteristic of parallel algorithms is speedup. We will use the traditional notion of speedup as the ratio of the time of sequential running to the time of solving the problem in the parallel mode. Fig.~\ref{fig4} and Fig.~\ref{fig5} show the graphs of speedup obtained by parallel running of iOpt and Optuna. We can clearly see that iOpt shows better speedup on a large number of processes compared to OPTUNA.

>>>>

%(рус.) Данная статья продолжает развитие подхода к параллельному решению задач липшицевой глобальной оптимизации, предложенного в работах [наши работы]. Новым элементом является сочетание способов распараллеливания и предложенной в [Усова] схемы работы с частично вычислимыми функциями. Основная часть публикации имеет следующую структуру. В Section 2 приведена математическая постановка решаемой задачи. Section 3 содержит описание параллельного алгоритма глобального поиска. В Section 4 описан способ решения задач со смешанными (дискретными и непрерывными) параметрами, а также обсуждается способ работы с функциями, определенными не во всей области поиска. Section 5 содержит результаты вычислительных экспериментов, демонстрирующих эффективность предлагаемого подхода при решении задачи оптимизации гиперпараметров методов прогнозирования значений временного ряда.
This paper continues to develop the approach to solving Lipschitz global optimization problems in parallel \cite{Barkalov2018,Gegrel2021,Strongin2018}. The new element is the combination of parallelization methods and the scheme for handling partially computable functions proposed in \cite{Usova2024}. The main part of the publication has the following structure.  Section \ref{sec:PS} presents the mathematical formulation of the problem to be solved. Section \ref{sec:PA} contains a description of the parallel global search algorithm. Section \ref{sec:MP} describes the method for solving problems with mixed (discrete and continuous) parameters, and also discusses the method for handling functions that are not defined in the entire search domain. Section \ref{sec:RCE} gives the results of computational experiments demonstrating the effectiveness of the proposed approach in solving the problem of optimizing the hyperparameters of time series forecasting methods.

@@

%(рус.) 6. Вычислить точку проведения нового испытаний x^{k+p+1} ∈ (x_(t-1),x_t ) в соответствии с формулами
\item Calculate the point of the new trial $x^{k+p+1} \in (x_{t-1}, x_t)$ according to the formulas
\begin{equation}\label{new_point_1} 
x^{k+p+1} = \frac{x_t+x_{t-1}}{2}, \text{ if } v(x_{t-1}) \neq v(x_t),
\end{equation}
\begin{equation}\label{new_point_2} 
x^{k+p+1} = \frac{x_t+x_{t-1}}{2}-\frac{\textit{sign}{(z_t-z_{t-1})}}{2r} \left[ \frac{|z_t-z_{t-1}|}{\mu} \right]^N,
\end{equation}
if $v(x_{t-1})=v(x_t)$.
%(рус.) Сразу после вычисления точки очередного испытания процесс-мастер добавляет ее в множество S и пересылает процессу-рабочему, который начинает проведение испытания в ней.
Immediately after calculating the point of the next trial, the master process adds it to the set $S$ and forwards it to the worker process, which starts a trial there.
\end{enumerate}

@@

%(рус.) Также требуется дополнить правило выбора точки нового испытания для обработки ситуаций, когда обе граничные точки имеют нулевой индекс:
It is also necessary to supplement the rule for selecting a new trial point to handle situations where both boundary points have a zero index:
\begin{equation}\label{new_point_1_2} 
x^{k+p+1} = \frac{x_t+x_{t-1}}{2}, \text{ if } v(x_{t-1}) \neq v(x_t) \text{ or } v(x_{t-1})=v(x_t)=0,
\end{equation}
\begin{equation}\label{new_point_2_2} 
x^{k+p+1} = \frac{x_t+x_{t-1}}{2}-\frac{\textit{sign}{(z_t-z_{t-1})}}{2r} \left[ \frac{|z_t-z_{t-1}|}{\mu} \right]^N,
\end{equation}
if $v(x_{t-1})=v(x_t) = 1$.

@@

%(рус.) Рассмотренные в статье методы реализованы в iOpt parallel solver, разрабатываемом в ННГУ им. Н.И. Лобачевского и предназначенном для решения многомерных задач глобальной оптимизации. 
The methods considered in this paper are implemented in the iOpt parallel solver \cite{iOptURL}, which is being developed at Lobachevsky State University of Nizhni Novgorod  for solving multidimensional global optimization problems.

%(рус.) В качестве модельной задачи нами была использована задача настройки гиперпараметров методов предсказания значений временного ряда (Time Series Forecasting). В качестве исходных данных был использован датасет montly beer prodaction [5.1], содержащий 476 valid data points. 
We used the task of hyperparameter tuning for time series forecasting methods as a model problem. The monthly beer production dataset \cite{MonthlyBeerDataset,MonthlyBeerArticle} containing 476 valid data points was used as input data.

@@

%(рус.) Построенный с использованием FEDOT pipeline включает в себя комбинацию из двух преобразований: lagged и cgru. Lagged-преобразование выполняется путем взятия значения переменной в предыдущий момент времени и включения его в модель в качестве признака в текущий момент времени. Для этого данные временного ряда сдвигаются на определенное количество шагов, которые называются лагом или временной задержкой. В качестве используемой нейронной сети будет выступать рекуррентная сеть (GRU), использующая свертку (CGRU).
The pipeline constructed using FEDOT involves a combination of two transformations: \textit{lagged} and \textit{cgru}, Fig.~\ref{fig2}. Lagged transformation is performed by taking the value of a variable at a previous point in time and incorporating it into the model as a feature at the current point in time. For this purpose, the time series data is shifted by a certain number of steps, which are called  a lag  or time delay. A recurrent network (GRU) using convolution (CGRU) will serve as the neural network.

@@

%(рус.) У cgru варьировались 4 дискретных параметра: cnn1_kernel_size  {4, 5}, cnn2_kernel_size  {5, 6}, cnn1_output_size  {32, 64}, cnn2_output_size  {32, 64}; и 2 непрерывных параметра: hidden_size  [20, 200], learning_rate  [0.0005, 0.005]. У остальных параметров были зафиксированы default values: batch_size = 64, num_epochs = 50, optimizer = 'adamw', loss = ‘mse’.
For CGRU, 4 discrete and 2 continuous parameters were varied:
\begin{itemize}
\item cnn1\_kernel\_size $\in \{4, 5\}$,
\item cnn2\_kernel\_size $\in \{5, 6\}$, 
\item cnn1\_output\_size $\in \{32, 64\}$, 
\item cnn2\_output\_size $\in \{32, 64\}$;
\item hidden\_size $\in [20, 200]$, 
\item learning\_rate $\in [0.0005, 0.005]$.
\end{itemize}

@@

%(рус.) Computations were conducted on the supercomputer “Lobachevsky” of the Nizhny Novgorod State University (operating system CentOS 7, control system SLURM, each supercomputer node had two processors Intel Xeon Silver 4310T CPU@2.30GHz (20 cores), 64Gb RAM). were used. Computations were performed using Python 3.9, FEDOT 0.7.5 [5.2, 5.2.1], iOpt  [20] and Optuna [5.3, 5.3.1].
Calculations were performed on the ``Lobachevsky'' supercomputer of the Nizhni Novgorod State University (CentOS 7 operating system, SLURM control system, each supercomputer node had two Intel Xeon Silver 4310T processors CPU@2.30GHz (20 cores), 64Gb RAM). Data analyses were carried out using Python 3.9, FEDOT 0.7.5 \cite{FEDOT}, iOpt 0.5.0 \cite{iOptURL} and Optuna 4.2.1 \cite{OptunaURL}.

@@

\begin{table}
\centering
\caption{Results of solving the problem using iOpt and Optuna.}\label{tab1}
\begin{tabular}{|l|c|c|c|c|}
\hline
\multirow{2}{*}{P} & \multicolumn{2}{c|}{iOpt} & \multicolumn{2}{c|}{Optuna} \\
\cline{2-5} 
 & Solving time & Final metric & Solving time & Final metric \\
\hline
1 & 6574.2 & 13.5 & 9554.1 & 13.4 \\
2 & 3579.2 & 13.9 & 5127.5 & 13.6 \\
4 & 1998.7 & 13.6 & 2598.6 & 13.2 \\
8 & 1331.7 & 13.2 & 1275.4 & 13.2 \\
16 & 846.6 & 13.3 & 1148.5 & 13.3 \\
20 & 699.9 & 13.9 & 1578.6 & 13.3 \\
\hline
\end{tabular}
\end{table}

%(рус.) Ключевой характеристикой параллельных алгоритмов является ускорение. Мы будем использовать традиционное понятие ускорения как отношение времени последовательного запуска ко времени решения задачи в параллельном режиме. На Fig.3 приведены графики ускорения, полученного при параллельном запуске iOpt и OPTUNA. Наглядно видно, что iOpt на большом числе процессов показывает лучшее ускорение по сравнению с OPTUNA.
The key characteristic of parallel algorithms is speedup. We will use the traditional notion of speedup as the ratio of the time of sequential running to the time of solving the problem in the parallel mode. Fig.~\ref{fig4} and Fig.~\ref{fig5} show the graphs of speedup and time obtained by parallel running of iOpt and Optuna. We can clearly see that iOpt shows better speedup on a large number of processes compared to Optuna.
